Algorithms for Estimating Trends in Global Temperature Volatility

### 2018-09-04 aaai-2018/khodadadi-mcdonald-aaai19.tex

%\pdfinfo{
%	/Title (Algorithms for Estimating Trends in Global Temperature Volatility)
	%/Author (Khodadadi and McDonald)
%	/Author (John Doe, Jane Doe)}

@@

\begin{abstract}
Trends in terrestrial temperature variability are perhaps more relavant for species
viability than trends in mean
temperature. In this paper, we develop methodology for estimating such
trends using multi-resolution climate data from polar orbiting weather
satellites. We derive two novel algorithms for computation that are
tailored for dense, gridded observations over both 
space and time. We evaluate our methods with a simulation constructed
to mimic these data's 
features and on a large, publicly available, global
temperature dataset with the goal of tracking trends in
cloud reflectance temperature variability.
\end{abstract}

@@

\textcolor{red}{How is this first part related to the rest of the paper?}

The amount of sunlight reflected from clouds is among the largest
sources of uncertainty in climate
prediction~\citep{BoucherRandall2013}. But climate models fail to
reproduce global cloud statistics and understanding the reasons for
this failure is a grand challenge of the World Climate Research
Programme~\citep{BonyStevens2015}.
By understanding how cloudliness changes over time, examining changes
in brightness temperature within various cloud types, and inferring
changes in cloud radiative effects, atmospheric scientists can better model climate
change. Modeling inter-satellite biases, satellite drift, and
seasonal and long-term climate phenomena like El Ni\~{n}o will lead to
a better understanding of climate
change~\citep{SchreierKahn2014,BaumWielicki1994,BaumWielicki1992,FreyAckerman1996}.

Numerous
studies have examined the overall impacts of clouds on climate
variability~\citep{MyersMechoso2018,GrisePolvani2013,BenderRamanathan2012},
but such investigations have been hampered by the lack of a suitable
dataset. Ideal data would have global coverage at high spatial
resolution, a long enough record to recover temporal trends, and be
multispectral~\citep{WielickiYoung2013}. The International Satellite
Cloud Climatology Project (ISCCP)
has been producing cloud property information from
limited spectral channels for over three
decades~\citep{RossowSchiffer1991}. These data have been used in 
various climate variability
studies~\citep[e.g.]{BenderRamanathan2012}, however, the utility of
these data for long-term study of climate variability has been
questioned~\citep{EvanHeidinger2007}.

ISCCP provides a relatively long record, but it only incorporates
radiances from limited visible and infrared channels from Advanced
Very High Resolution Radiometer (AVHRR) instruments. Ongoing work seeks to create a
more spectrally-detailed dataset which can avoid the above issues by
combining radiance data from AVHRR imagers with readings from High-resolution
Infrared Radiation Sounders on board legacy
satellites~\citep{StatenKahn2016,SchreierKahn2010,KahnFishbein2007}. In
anticipation of this new dataset, our work develops novel methodology
for examining the trends in variability of climate data across space
and time.

@@

Nevertheless, research examining trends in the volatility of
spatio-temporal climate data is scarce. \citet{hansen_perception_2012} studied the change in
the standard deviation (SD) of the surface temperature in the NASA
Goddard Institute for Space Studies gridded temperature data set by
examining the empirical SD at each spatial location relative to that
location's SD over a base period, and showed
that these estimates are increasing.
\citet{huntingford_no_2013} took a similar
approach in analyzing the ERA-40 data set. Their results showed that there
still is an increase in the SDs from 1958-1970 to 1991-2001, but this
is much less than what is obtained from the method used in
\cite{hansen_perception_2012}. The authors also computed the
time-evolving global SD from the de-trended time-series at each
position, which suggests that the global SD has been
stable.

These and other related research, e.g.,
\citet{rhines_frequent_2013}, have several shortcomings. First, no
statistical analysis has been performed to examine if the changes in
the SD are statistically significant. Second, the methodologies for
computing the SDs are highly sensitive to the choice of base period. Third,
and most importantly, temporal and spatial correlations between the
observations are completely ignored.

@@

The main contribution of this work is to develop a new methodology for
detecting the trend in the volatility of spatio-temporal data. In this
methodology, the variance at each position and time is considered as
a hidden variable. The values of these hidden variables
are then estimated by maximizing the likelihood of the observed
data. Following~\cite{Tibshirani2014}, we penalize the
differences between the estimated variances which
are temporally and spatially ``close'', resulting
in a generalized LASSO problem. However, in our application, the
dimension of this optimization problem is massive, and so the
standard solvers are inadequate. We develop two algorithms
which are computationally feasible. In the first method, we adopt an optimization technique
called alternating direction method of multipliers
(ADMM, \citealp{boyd_distributed_2011}), to divide the total problem into
several sub-problems of much lower dimension and show how the total
problem can be solved by iteratively solving these sub-problems. The
second method, called the \textit{linearized ADMM algorithm}
\cite{parikh_proximal_2014} solves the main problem by iteratively
solving a linearized version of it. We will compare the benefits of
each method.

Our main contributions are as follows:
\begin{enumerate}
\item We propose a model for nonparametric variance estimation for a
  spatio-temporal process and discuss the relationship between our
  methods and those existing in the machine learning literature (\autoref{sec:ell_1-trend-filt}).
\item We derive two alternating direction method of multiplier
  algorithms (ADMM) to fit our estimator when applied to 
  very large data (\autoref{sec:prop-optim-meth}). We give situations
  under which each algorithm is most likely to be useful. Open-source
  Python code is available on {\tt github}.
\item We illustrate our methods on a large, publicly available, global temperature dataset
  with the goal of tracking world-wide trends in variance as well as a
  simulation constructed to mimic these data's features
  (\autoref{sec:empirical-evaluation}). 
\end{enumerate}

@@

\citet{KimKoh2009} proposed $\ell_1$-trend filtering as a
method for estimating a smooth, time-varying trend. It is formulated
as the optimization problem 
% : a time-series $y_t$, $t=1,...,T$ is observed and we believe that
% it consists of a smooth trend and a stochastic component and the
% goal is to estimate the trend. The estimated trend is desired to be
% smooth, but at the same time, we seek small estimated residuals
% (stochastic component). These two objectives cannot be achieved
% simultaneously and so a trade-off between them should be made. In
% $\ell_1$-trend filtering, this trade-off is formulated as the
% following convex optimization problem: 
$
\min_{\beta} \frac{1}{2} \sum_{t=1}^{T} (y_t-\beta_t)^2+\lambda
\sum_{t=1}^{T-2} \left|\beta_t-2\beta_{t+1}+\beta_{t+2}\right| 
$
or equivalently:
\begin{equation}
\min_{\beta} \frac{1}{2} \norm{ y-\beta }_2^2+\lambda \norm{ D_t \beta}_1
\label{eq:l1tf}
\end{equation}
 where $y=\{y_t\}_{t=1}^T$ is an observed time-series, $\beta\in \mathbb{R}^T$ is the smooth trend,
 $D_t$ is a $(T-2)\times T$ matrix, and $\lambda$ is a tuning parameter
 which balances fidelity to the data (small errors in the first term)
 with a desire for smoothness.  
% The \textit{penalty matrix} is:
% \begin{equation}
% D=\begin{bmatrix}
% 1 & -2 & 1 &  &  &  &\\ 
% & 1 & -2 & 1 &  &  &\\ 
% &  & \ddots & \ddots & \ddots  &  &\\ 
% &  & & 1 & -2 & 1 &  \\ 
% &  &  &   & 1 & -2 & 1 
% \end{bmatrix}
% \label{eq:d_matrix}
% \end{equation}
%The first term in \eqref{eq:l1tf} penalizes large residuals while the second term encourages smooth estimated $\beta$.
In Appendix A, we specify the form of the matrix $D_t$ which results in piecewise
linear estimated $\beta$. \citet{KimKoh2009} proposed a specialized primal-dual
interior point (PDIP) algorithm for solving \eqref{eq:l1tf}. From a
statistical perspective, \eqref{eq:l1tf} can be viewed as a constrained maximum
likelihood problem with independent observations from a normal
distribution with common variance, $y_t \sim \mbox{N}(\beta_t,
\sigma^2)$, subject to a piecewise linear constraint on
$\beta$. Alternatively, solutions to \eqref{eq:l1tf} are equivalent to
maximum a posteriori Bayesian estimators based on Gaussian likelihood
with a special Laplace prior distribution on $\beta$. Note that the
structure of the estimator is determined by the penalty function
$\lambda\norm{D_t\beta}_1$ rather than any parametric trend
assumptions---autoregressive, moving average, sinusoidal seasonal
component, etc. The resulting trend is therefore essentially
nonparametric in the same way that splines are. In fact,
using squared $\ell_2$ norm as the penalty instead of $\ell_1$ results exactly in
regression splines.

@@

Inspired by the $\ell_1$-trend filtering algorithm, we propose a
non-parametric model for estimating the variance of a time-series. To
this end, we assume that at each time step $t$, there is a parameter
$h_t$ such that the observations $y_t$
are independent normal variables with zero mean and variance
$\exp(h_t)$. The negative log-likelihood of the observed data in this
model is $l(y\given h) \propto -\sum_{t=1}^T h_t - y_t^2e^{-h_t}$. Crucially,
we assume that the parameters $h_t$ vary smoothly. To impose
this assumption, we estimate $h_t$ by solving the penalized, negative
log-likelihood: 
\begin{equation}
\min_h -l(y\given h)+\lambda \norm{ D_th }_1
\label{eq:l1tf_var}
\end{equation}
where $D_t$ has the same structure as above.

As with~\eqref{eq:l1tf}, one can solve~\eqref{eq:l1tf_var} using the
PDIP algorithm (as in, e.g.,
\texttt{cvxopt}, \citealp{andersen_cvxopt:_2013}). In each iteration of
PDIP we need to compute a search direction by taking a Newton
step on a system of nonlinear equations. Due to space limitations, we
defer details to Appendix A in the Supplement, where we show how to
derive the dual of this optimization problem and compute the first and
second derivatives of the dual objective function.

@@

The method in the previous section can be used to estimate the
variance of a single time-series. In this section, we extend this
method to the estimation of the variance of spatio-temporal data.

At a specific time $t$, the data are measured on a grid of points with
$n_r$ rows and $n_c$ columns for a total of $S=n_r\times n_c$ spatial
locations. Let $y_{ijt}$ denote the value of the 
observation at time $t$ on the $i^\text{th}$ row and $j^\text{th}$
column of the grid, and $h_{ijt}$ denote the corresponding hidden
variable. We seek to impose both temporal and spatial smoothness
constraints on the hidden variables. Specifically, we seek a solution
for $h$ which is piecewise linear in time and piecewise constant in
space (although higher-order smoothness can be imposed with minimal
alterations to the methodology). We achieve this goal
by solving the following optimization problem: 
{\scriptsize
\begin{equation}
\label{eq:l1tf_var_st}.
\begin{aligned}
&\min_h \sum_{i,j,t}h_{ijt}+y_{ijt}^2e^{-h_{ijt}} +\lambda_t
\sum_{i,j} \sum_{t=1}^{T-2}
\left|h_{ijt}-2h_{ij(t+1)}+h_{ij(t+2)}\right|\\ 
&+\lambda_s \sum_{t,j} \sum_{i=1}^{n_r-1} \left|h_{ijt}-h_{(i+1)jt}\right|
+\lambda_s \sum_{t,i} \sum_{j=1}^{n_c-1} \left|h_{ijt}-h_{i(j+1)t}\right|
\end{aligned}
\end{equation}
}%

The first term in the objective is proportional to the negative
log-likelihood, the second is the temporal penalty for the
time-series at each location $(i,j)$, while the third and fourth,
penalize the difference between the estimated variance of two
vertically and horizontally adjacent points, respectively. The spatial
component of this
penalty is a special case of trend filtering on
graphs~\cite{WangSharpnack2016} which penalizes the difference between
the estimated values of the signal on the connected nodes (though the
likelihood is different). As before,
we can write \eqref{eq:l1tf_var_st} in matrix form where $h$ is an
$T\times S$ vector and $D_t$ is replaced by $D \in
\mathbb{R}^{(N_t+N_s) \times (T \cdot S)	}$, where $N_t=S \cdot
(T-2)$ and $N_s=T \cdot (2n_rn_c-n_r)$ are the number of temporal and
spatial constraints, respectively\footnote{$N_s$ is obtained by
  counting the number of unique constraints at each location and at
  all times.}. The exact form of this matrix in clarified in Appendix
A in the Supplement. Then, as we have two different tuning parameters
for the temporal and spatial components, we write $\Lambda
=\left[\lambda_1\one_{N_t}^\top,\;
  \lambda_2\one_{N_s}^\top\right]^\top$ leading
to:\footnote{Throughout the paper, we use $|x|$ for both scalars and
  vectors. For vectors we use this to denote a vector obtained by
  taking the absolute value of each entry of $x$.}  
% let $h$ be a
% vector whose first $T$ entries are $h_{11t}$ for $t=1,...,T$, its next
% $T$ entries are $h_{21t}$ and so on, then the optimization problem in
% the matrix form is as
\begin{equation}
\min_h -l(y\given h)+ \Lambda^\top | Dh |.
\label{eq:l1tf_var_st_mat}
\end{equation}

@@

\attn{I've moved this here rather than above to focus the intro more
  on the data.}

@@

Our methodology is similar in flavor to~\cite{HallacPark2017} or
related work in~\citep{GibberdNelson2017,MontiHellyer2014}, but with
several fundamental differences. These papers aim to
discover the time-varying structure of a network. To achieve this
goal, they use Gaussian likelihood with unknown precision matrix and
introduce penalty terms which (1) encourage sparsity among the
off-diagonal elements and (2) discourage changes in the estimated inverse
covariance matrix from one time-step to the next.
Our goal in the present work is to detect the temporal trend in the variance of each
point in the network, but the network is known (corresponding to the grid
over the earth). The variance of each point, however changes. To modify the
objective function in~\cite[Eq.\ 2]{HallacPark2017}, we would
enforce complete sparsity on the off-diagonal elements (since
they are not estimated) and add a new penalty to enforce spatial
behavior across the diagonal elements. Thus, 
\eqref{eq:l1tf_var_st_mat} is not simply a special case of these
existing methods.

@@

For a spatial grid of size $S$ and $T$ time steps, $D$ in equation \ref{eq:l1tf_var_st_mat} will have
$3Tn_rn_c-2n_rn_c-Tn_r$ rows and $S \cdot T$ columns. For a $1^\circ\times
1^\circ$ grid over the entire northern hemisphere and daily data over
10 years, we have $S=90\times 360\approx 32,000$ spatial locations and
$T=3650$ time points, and so $D$ has
approximately $10^8$ columns and $10^8$ rows. In each step of the PDIP
algorithm, we need to solve a linear system of equations which
depends on $D^\top D$ (see appendix A and B). Therefore,
applying the PDIP directly is infeasible for our data.\footnote{We
  note that $D$ is a highly structured, sparse matrix, but, unlike
  trend filtering alone, it is not banded. We are unaware of general
  linear algebra techniques for inverting such matrix, despite our
  best efforts.}

@@

Assume it is possible to define a set of local variables
$x_i \in \mathbb{R}^{n_i}$ such that $f(h)=\sum_i f_i(x_i)$, where
each $x_i$ is a subset of the global variable $h$. More specifically,
each entry of the local variables corresponds to an entry of the
global variable. Therefore we can define a mapping $\mathscr{G}(i,j)$
from the local variable indices into the global variable indices:
$k=\mathscr{G}(i,j)$ means that the $j^\text{th}$ entry of $x_i$ is
$h_k$ (or $(x_i)_j=h_k$). For ease of notation, define $\tilde{h}_i
\in \mathbb{R}^{n_i}$ as $(\tilde{h}_i)_j=h_{\mathscr{G}(i,j)}$. Then,
the original optimization problem is equivalent to the following
problem:  
\begin{equation}
\begin{aligned}
&\min_{\{x_1,...,x_N  \}} \sum_i f_i(x_i) &
s.t. &\quad \tilde{h}_i=x_i.
\end{aligned}
\label{eq:consADMM}
\end{equation}
It is important to note that each entry of the global variable may
correspond to several entries of the local variables and so the
constraints $\tilde{h}_i=x_i$ enforce the consensus between the local
variables corresponding to the same global variable.  
The \textit{augmented Lagrangian} corresponding to 
\eqref{eq:consADMM} is $L_\rho(x,h,y)=\sum_i
\big(f_i(x_i)+u_i^\top(x_i-\tilde{h}_i) + (\rho/2) \lVert
x_i-\tilde{h}_i \lVert_2^2 \big)$. Now, we can apply ADMM to $L_\rho$. This results in solving $N$ independent optimization problems followed by a step to achieve consensus among the solutions in each iteration.
% which results in the following ADMM updates:
% \begin{align}
% x_i&\leftarrow\argmin_{x_i} f_i(x_i) + (u_i)^\top x_i +
%      (\rho/2)  \norm{ x_i-\tilde{h}_i }_2^2\\
% h_k&\leftarrow (1/S_k)\sum_{\mathscr{G}(i,j)=k} (x_i)_j \label{eq:ADMM_steps}\\
% u_i&\leftarrow u_i + \rho (x_i-\tilde{h}_i).
% \end{align}
% \attn{If you hate the left arrow, that's fine. But change it
%   everywhere, and use $=$. The $:=$ notation means that ``thing on
%   left is defined to be thing on right'' and isn't really appropriate here.}
% Here, $S_k$ is the number of local variable entries that correspond to
% $h_k$, and $u_i$ are the Lagrange multipliers.  
To solve the optimization problem \eqref{eq:l1tf_var_st_mat} using
this method, we need to address two 
questions: first, how to choose the 
local variables $x_i$, and second, how to the update them.
\begin{figure}[tb]
  \centering
  \includegraphics[width=.6\columnwidth]{Figures/data_cube}
  \caption{The cube represents the global variable $h$ in space and
    time. The sub-cubes specified by the white lines are
    $x_i$.}
  \label{fig:data_cube}
\end{figure}

@@

\paragraph{Convergence}

As shown in \autoref{alg:linADMM}, we evaluated convergence using
$\epsilon=0.001\%$ of the MSE of the data.
Our simulation experiments showed that the
convergence speed depends on the value of $\lambda_t$ and
$\lambda_s$. Furthermore, using the solution obtained for smaller values
of these parameters as a warm start for the larger values, the
converges speed improves.

>>>>

\pdfinfo{
  /Title (Algorithms for Estimating Trends in Global Temperature Volatility)
  % /Author (Khodadadi and McDonald)
  /Author (John Doe, Jane Doe)}

@@

\begin{abstract}
Trends in terrestrial temperature variability are perhaps more relavant for species
viability than trends in mean
temperature. In this paper, we develop methodology for estimating such
trends using multi-resolution climate data from polar orbiting weather
satellites. We derive two novel algorithms for computation that are
tailored for dense, gridded observations over both 
space and time. We evaluate our methods with a simulation constructed
to mimic these data's 
features and on a large, publicly available, global
temperature dataset with the eventual goal of tracking trends in
cloud reflectance temperature variability.
\end{abstract}

@@

The amount of sunlight reflected from clouds is among the largest
sources of uncertainty in climate
prediction~\citep{BoucherRandall2013}. But climate models fail to
reproduce global cloud statistics, and understanding the reasons for
this failure is a grand challenge of the World Climate Research
Programme~\citep{BonyStevens2015}.
% By understanding how cloudliness changes over time, examining changes
% in brightness temperature within various cloud types, and inferring
% changes in cloud radiative effects, atmospheric scientists can better model climate
% change. Modeling inter-satellite biases, satellite drift, and
% seasonal and long-term climate phenomena like El Ni\~{n}o will lead to
% a better understanding of climate
% change~\citep{SchreierKahn2014,BaumWielicki1994,BaumWielicki1992,FreyAckerman1996}.
While numerous
studies have examined the overall impacts of clouds on climate
variability~\citep{MyersMechoso2018,GrisePolvani2013,BenderRamanathan2012},
such investigations have been hampered by the lack of a suitable
dataset. Ideal data would have global coverage at high spatial
resolution, a long enough record to recover temporal trends, and be
multispectral~\citep{WielickiYoung2013}. 
% The International Satellite
% Cloud Climatology Project (ISCCP)
% has been producing cloud property information from
% limited spectral channels for over three
% decades~\citep{RossowSchiffer1991}. These data have been used in 
% various climate variability
% studies~\citep[e.g.]{BenderRamanathan2012}, however, the utility of
% these data for long-term study of climate variability has been
% questioned~\citep{EvanHeidinger2007}. 
% ISCCP provides a relatively long record, but it only incorporates
% radiances from limited visible and infrared channels from Advanced
% Very High Resolution Radiometer (AVHRR) instruments. 
Current work~\cite{StatenKahn2016,SchreierKahn2010,KahnFishbein2007} seeks to create a
spectrally-detailed dataset by %which can avoid the above issues by
combining radiance data from Advanced Very High Resolution Radiometer
imagers with readings from High-resolution 
Infrared Radiation Sounders on board legacy
satellites. 
In anticipation of this new dataset, our work develops novel methodology
for examining the trends in variability of climate data across space
and time.

@@

Nevertheless, research examining trends in the volatility of
spatio-temporal climate data is scarce. \citet{hansen_perception_2012} studied the change in
the standard deviation (SD) of the surface temperature in the NASA
Goddard Institute for Space Studies gridded temperature data set by
examining the empirical SD at each spatial location relative to that
location's SD over a base period, and showed
that these estimates are increasing.
\citet{huntingford_no_2013} took a similar
approach in analyzing the ERA-40 data set. They argued that, while there
is an increase in the SDs from 1958-1970 to 1991-2001, it
is much smaller than found in
\cite{hansen_perception_2012}. \citet{huntingford_no_2013} also computed the
time-evolving global SD from the de-trended time-series at each
position and argued that the global SD has been
stable.

These and other related work  
\cite[e.g.,]{rhines_frequent_2013} have several shortcomings. First, no
statistical analysis has been performed to examine if the changes in
the SD are statistically significant. Second, the methodologies for
computing the SDs are highly sensitive to the choice of base period. Third,
and most importantly, temporal and spatial correlations between 
observations are completely ignored.

@@

The main contribution of this work is to develop a new methodology for
detecting the trend in the volatility of spatio-temporal data. In this
methodology, the variance at each position and time are estimated by
maximizing penalized likelihood
maximization. Following~\cite{Tibshirani2014}, we penalize the 
differences between the estimated variances which
are temporally and spatially ``close'', resulting
in a generalized LASSO problem. However, in our application, the
dimension of this optimization problem is massive, so the
standard solvers are inadequate.

We develop two algorithms
which are computationally feasible on extremely large data. In the
first method, we adopt an optimization technique 
called alternating direction method of multipliers
(ADMM, \citealp{boyd_distributed_2011}), to divide the total problem into
several sub-problems of much lower dimension and show how the total
problem can be solved by iteratively solving these sub-problems. The
second method, called linearized ADMM
\cite{parikh_proximal_2014}, solves the main problem by iteratively
solving a linearized version of it. We will compare the benefits of
each method.

Our main contributions are as follows:
\begin{enumerate}
\item We propose a model for nonparametric variance estimation for a
  spatio-temporal process and discuss the relationship between our
  methods and those existing in the machine learning literature (\autoref{sec:ell_1-trend-filt}).
\item We derive two alternating direction method of multiplier
  algorithms (ADMM) to fit our estimator when applied to 
  very large data (\autoref{sec:prop-optim-meth}). We give situations
  under which each algorithm is most likely to be useful. Open-source
  Python code is available on {\tt github}.
\item Because the construction of satellite-based datasets is ongoing
  and currently proprietary, we illustrate our methods on a large,
  publicly available, global temperature dataset.
  The goal is to demonstrate the feasibility of these methods for tracking
  world-wide trends in variance in standard atmospheric data and a
  simulation constructed to mimic these data's features
  (\autoref{sec:empirical-evaluation}). 
\end{enumerate}

@@

\citet{KimKoh2009} proposed $\ell_1$-trend filtering as a
method for estimating a smooth, time-varying trend. It is formulated
as the optimization problem 
% : a time-series $y_t$, $t=1,...,T$ is observed and we believe that
% it consists of a smooth trend and a stochastic component and the
% goal is to estimate the trend. The estimated trend is desired to be
% smooth, but at the same time, we seek small estimated residuals
% (stochastic component). These two objectives cannot be achieved
% simultaneously and so a trade-off between them should be made. In
% $\ell_1$-trend filtering, this trade-off is formulated as the
% following convex optimization problem: 
$$
\min_{\beta} \frac{1}{2} \sum_{t=1}^{T} (y_t-\beta_t)^2+\lambda
\sum_{t=1}^{T-2} \left|\beta_t-2\beta_{t+1}+\beta_{t+2}\right| 
$$
or equivalently:
\begin{equation}
\min_{\beta} \frac{1}{2} \norm{ y-\beta }_2^2+\lambda \norm{ D_t \beta}_1
\label{eq:l1tf}
\end{equation}
 where $y=\{y_t\}_{t=1}^T$ is an observed time-series, $\beta\in \mathbb{R}^T$ is the smooth trend,
 $D_t$ is a $(T-2)\times T$ matrix, and $\lambda$ is a tuning parameter
 which balances fidelity to the data (small errors in the first term)
 with a desire for smoothness.  
% The \textit{penalty matrix} is:
% \begin{equation}
% D=\begin{bmatrix}
% 1 & -2 & 1 &  &  &  &\\ 
% & 1 & -2 & 1 &  &  &\\ 
% &  & \ddots & \ddots & \ddots  &  &\\ 
% &  & & 1 & -2 & 1 &  \\ 
% &  &  &   & 1 & -2 & 1 
% \end{bmatrix}
% \label{eq:d_matrix}
% \end{equation}
%The first term in \eqref{eq:l1tf} penalizes large residuals while the second term encourages smooth estimated $\beta$.
% In Appendix A, we specify the form of the matrix $D$ which results in piecewise
% linear estimated $\beta$. 
\citet{KimKoh2009} proposed a specialized primal-dual
interior point (PDIP) algorithm for solving \eqref{eq:l1tf}. From a
statistical perspective, \eqref{eq:l1tf} can be viewed as a constrained maximum
likelihood problem with independent observations from a normal
distribution with common variance, $y_t \sim \mbox{N}(\beta_t,
\sigma^2)$, subject to a piecewise linear constraint on
$\beta$. Alternatively, solutions to \eqref{eq:l1tf} are %equivalent to
maximum a posteriori Bayesian estimators based on Gaussian likelihood
with a special Laplace prior distribution on $\beta$. Note that the
structure of the estimator is determined by the penalty function
$\lambda\norm{D_t\beta}_1$ rather than any parametric trend
assumptions---autoregressive, moving average, sinusoidal seasonal
component, etc. The resulting trend is therefore essentially
nonparametric in the same way that splines are nonparametric. In fact,
using squared $\ell_2$ norm as the penalty instead of $\ell_1$ results exactly in
regression splines.

@@

Inspired by the $\ell_1$-trend filtering algorithm, we propose a
non-parametric model for estimating the variance of a time-series. To
this end, we assume that at each time step $t$, there is a parameter
$h_t$ such that the observations $y_t$
are independent normal variables with zero mean and variance
$\exp(h_t)$. The negative log-likelihood of the observed data in this
model is $l(y\given h) \propto -\sum_{t=1}^T h_t - y_t^2e^{-h_t}$. Crucially,
we assume that the parameters $h_t$ vary smoothly and estimate them by
minimizing the penalized, negative
log-likelihood:  
\begin{equation}
\min_h -l(y\given h)+\lambda \norm{ D_th }_1
\label{eq:l1tf_var}
\end{equation}
where $D_t$ has the same structure as above.

As with~\eqref{eq:l1tf}, one can solve~\eqref{eq:l1tf_var} using the
PDIP algorithm (as in, e.g.,
\texttt{cvxopt}, \citealp{andersen_cvxopt:_2013}). In each iteration of
PDIP we need to compute a search direction by taking a Newton
step on a system of nonlinear equations. For completeness, we
provide the details in Appendix A of the Supplement, where we show how to
derive the dual of this optimization problem and compute the first and
second derivatives of the dual objective function.

@@

The method in the previous section can be used to estimate the
variance of a single time-series. Here we extend this
method to the case of spatio-temporal data.

At a specific time $t$, the data are measured on a grid of points with
$n_r$ rows and $n_c$ columns for a total of $S=n_r\times n_c$ spatial
locations. Let $y_{ijt}$ denote the value of the 
observation at time $t$ on the $i^\text{th}$ row and $j^\text{th}$
column of the grid, and $h_{ijt}$ denote the corresponding
parameter. We seek to impose both temporal and spatial smoothness 
constraints on the hidden variables. Specifically, we seek a solution
for $h$ which is piecewise linear in time and piecewise constant in
space (although higher-order smoothness can be imposed with minimal
alterations to the methodology). We achieve this goal
by solving the following optimization problem: 
{\scriptsize
\begin{equation}
\label{eq:l1tf_var_st}.
\begin{aligned}
&\min_h \sum_{i,j,t}h_{ijt}+y_{ijt}^2e^{-h_{ijt}} +\lambda_t
\sum_{i,j} \sum_{t=1}^{T-2}
\left|h_{ijt}-2h_{ij(t+1)}+h_{ij(t+2)}\right|\\ 
&+\lambda_s \sum_{t,j} \sum_{i=1}^{n_r-1} \left|h_{ijt}-h_{(i+1)jt}\right|
+\lambda_s \sum_{t,i} \sum_{j=1}^{n_c-1} \left|h_{ijt}-h_{i(j+1)t}\right|
\end{aligned}
\end{equation}
}%

The first term in the objective is proportional to the negative
log-likelihood, the second is the temporal penalty for the
time-series at each location $(i,j)$, while the third and fourth,
penalize the difference between the estimated variance of two
vertically and horizontally adjacent points, respectively. The spatial
component of this
penalty is a special case of trend filtering on
graphs~\cite{WangSharpnack2016} which penalizes the difference between
the estimated values of the signal on the connected nodes (though the
likelihood is different). As before,
we can write \eqref{eq:l1tf_var_st} in matrix form where $h$ is a
vector of length
$TS$ and $D_t$ is replaced by $D \in
\mathbb{R}^{(N_t+N_s) \times (T \cdot S)	}$, where $N_t=S \cdot
(T-2)$ and $N_s=T \cdot (2n_rn_c-n_r)$ are the number of temporal and
spatial constraints, respectively. 
% The exact form of this matrix in clarified in Appendix
% B in the Supplement. 
Then, as we have two different tuning parameters
for the temporal and spatial components, we write $\Lambda
=\left[\lambda_t\one_{N_t}^\top,\;
  \lambda_s\one_{N_s}^\top\right]^\top$ leading
to:\footnote{Throughout the paper, we use $|x|$ for both scalars and
  vectors. For vectors we use this to denote a vector obtained by
  taking the absolute value of each entry of $x$.}  
% let $h$ be a
% vector whose first $T$ entries are $h_{11t}$ for $t=1,...,T$, its next
% $T$ entries are $h_{21t}$ and so on, then the optimization problem in
% the matrix form is as
\begin{equation}
\min_h -l(y\given h)+ \Lambda^\top | Dh |.
\label{eq:l1tf_var_st_mat}
\end{equation}

@@

@@

Our methodology is similar in flavor to~\cite{HallacPark2017} or
related work in~\citep{GibberdNelson2017,MontiHellyer2014}, but with
several fundamental differences. These papers aim to
discover the time-varying structure of a network. To achieve this
goal, they use Gaussian likelihood with unknown precision matrix and
introduce penalty terms which (1) encourage sparsity among the
off-diagonal elements and (2) discourage changes in the estimated inverse
covariance matrix from one time-step to the next.
Our goal in the present work is to detect the temporal trend in the variance of each
point in the network, but the network is known (corresponding to the grid
over the earth). To modify the
objective function in~\cite[Eq.\ 2]{HallacPark2017}, we would
enforce complete sparsity on the off-diagonal elements (since
they are not estimated) and add a new penalty to enforce spatial
behavior across the diagonal elements. Thus, 
\eqref{eq:l1tf_var_st_mat} is not simply a special case of these
existing methods. Finally, these papers examine networks with hundreds
of nodes and dozens to hundreds of time points. As discussed next, our data are
significantly larger than these networks and attempting to estimate a
full covariance would be prohibitive, were it required.

@@

For a spatial grid of size $S$ and $T$ time steps, $D$ in equation \ref{eq:l1tf_var_st_mat} will have
$3Tn_rn_c-2n_rn_c-Tn_r$ rows and $S \cdot T$ columns. For a $1^\circ\times
1^\circ$ grid over the entire northern hemisphere and daily data over
10 years, we have $S=90\times 360\approx 32,000$ spatial locations and
$T=3650$ time points, so $D$ has
approximately $10^8$ columns and $10^8$ rows. In principal, we could
solve~\eqref{eq:l1tf_var_st_mat} using PDIP as before, however, each iteration
requires solving a linear system of equations which
depends on $D^\top D$. Therefore,
applying the PDIP directly is infeasible for our data.\footnote{We
  note that $D$ is a highly structured, sparse matrix, but, unlike
  trend filtering alone, it is not banded. We are unaware of general
  linear algebra techniques for inverting such matrix, despite our
  best efforts to find them.}

@@

Assume it is possible to define a set of local variables
$x_i \in \mathbb{R}^{n_i}$ such that $f(h)=\sum_i f_i(x_i)$, where
each $x_i$ is a subset of the global variable $h$. More specifically,
each entry of the local variables corresponds to an entry of the
global variable. Therefore we can define a mapping $\mathscr{G}(i,j)$
from the local variable indices into the global variable indices:
$k=\mathscr{G}(i,j)$ means that the $j^\text{th}$ entry of $x_i$ is
$h_k$ (or $(x_i)_j=h_k$). For ease of notation, define $\tilde{h}_i
\in \mathbb{R}^{n_i}$ as $(\tilde{h}_i)_j=h_{\mathscr{G}(i,j)}$. Then,
the original optimization problem is equivalent to the following
problem:  
\begin{equation}
\begin{aligned}
&\min_{\{x_1,...,x_N  \}} \sum_{i=1}^N f_i(x_i) &
s.t. &\quad \tilde{h}_i=x_i.
\end{aligned}
\label{eq:consADMM}
\end{equation}
It is important to note that each entry of the global variable may
correspond to several entries of the local variables and so the
constraints $\tilde{h}_i=x_i$ enforce the consensus between the local
variables corresponding to the same global variable.  
The augmented Lagrangian corresponding to 
\eqref{eq:consADMM} is $L_\rho(x,h,y)=\sum_i
\big(f_i(x_i)+u_i^\top(x_i-\tilde{h}_i) + (\rho/2) \lVert
x_i-\tilde{h}_i \lVert_2^2 \big)$. Now, we can apply ADMM to
$L_\rho$. This results in solving $N$ independent optimization
problems followed by a step to achieve consensus among the solutions
in each iteration. 
% which results in the following ADMM updates:
% \begin{align}
% x_i&\leftarrow\argmin_{x_i} f_i(x_i) + (u_i)^\top x_i +
%      (\rho/2)  \norm{ x_i-\tilde{h}_i }_2^2\\
% h_k&\leftarrow (1/S_k)\sum_{\mathscr{G}(i,j)=k} (x_i)_j \label{eq:ADMM_steps}\\
% u_i&\leftarrow u_i + \rho (x_i-\tilde{h}_i).
% \end{align}
% \attn{If you hate the left arrow, that's fine. But change it
%   everywhere, and use $=$. The $:=$ notation means that ``thing on
%   left is defined to be thing on right'' and isn't really appropriate here.}
% Here, $S_k$ is the number of local variable entries that correspond to
% $h_k$, and $u_i$ are the Lagrange multipliers.  
To solve the optimization problem \eqref{eq:l1tf_var_st_mat} using
this method, we need to address two 
questions: first, how to choose the 
local variables $x_i$, and second, how to the update them.
\begin{figure}[tb]
  \centering
  \includegraphics[width=.6\columnwidth]{Figures/data_cube}
  \caption{The cube represents the global variable $h$ in space and
    time. The sub-cubes specified by the white lines are
    $x_i$.}
  \label{fig:data_cube}
\end{figure}

@@

%\paragraph{Convergence and warm starts}

As shown in \autoref{alg:linADMM}, we evaluated convergence using
$\epsilon=0.001\%$ of the MSE of the data.
Our simulation experiments indicated that the
convergence speed depends on the value of $\lambda_t$ and
$\lambda_s$. For the temperature data, we used the solutions obtained for smaller values
of these parameters as a warm start for the larger values resulting in
faster convergence.
